Modelling fine-scale dive behavior with hidden Markov models

### 2020-06-25 Report/Sections/carhmm_ou.tex
Update Report

Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then depend upon the time step between observations. A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal. Namely, a state-switching Ornstein-Uhlenbeck process $Y$ is the solution to the following stochastic differential equation:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \cite{Michelot:2019}:
%
\begin{align}
Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)}^2}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
\label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$. However, most continuous time models require MCMC algorithms to perform inference over and as a result are not easily incorporated into the HHMM structure.

@@

Although most continuous time models are difficult to incorporate into an HHMM, we show in theorem (1) that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in equation (\ref{eqn:OU_sol}). This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.

\begin{theorem}{Theorem 1.}{}%
If:
\begin{enumerate}
    \item The hidden behavioural process $X$ from (Eq. \ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
\end{enumerate}
then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, where:

\begin{align}
\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{2(i)} = \frac{\omega^{(i)}^2}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t}), \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
\end{align}

>>>>

Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then have to depend upon the time step between observations. A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal. Namely, a state-switching Ornstein-Uhlenbeck process $Y$ is the solution to the following stochastic differential equation:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \cite{Michelot:2019}:
%
\begin{align}
Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)}^2}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
\label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$. However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.

@@

Although most continuous time models are difficult to incorporate into an HHMM, we show in (thrm 1) that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in (eqn \ref{eqn:OU_sol}). This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.

\begin{theorem}{Theorem 1.}{}%
If:
\begin{enumerate}
    \item The hidden behavioural process $X$ from (Eq. \ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
\end{enumerate}
then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

\begin{align}
\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
\end{align}

### 2020-06-25 Report/Sections/intro.tex
Update Report

>>>>

The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling. In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. This results in a vast amount of data which often exhibits many different simultaneous behavioral processes occurring at different time scales.

One solution to this issue is to use a hierarchical hidden Markov model (HHMM). HHMMs model the entire system as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process. One nice property of HHMMs is that its likelihood is relatively easy to compute, facilitating fast maximum likelihood estimates for its associated parameters.

At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by a traditional HMMs. To address this issue, it is possible to model small-scale animal behavior as the solution to some stochastic differential equation, but these methods tend to be computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC).

This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. We bridge the gap between discrete hidden Markov models and continuous-time stochastic process models by showing that the two are equivalent under certain conditions. In addition, we extract features from highly structured sub-dive behaviors using signal processing techniques. These features otherwise could not be modeled with a simple HMM. Finally, we apply our method to dive data collected from two Northern resident killer whales off the coast of British Columbia, Canada.

### 2020-06-25 Report/Sections/model.tex
Update Report

\subsection{Short-time Fourier Transform on Fine-scale process}

One issue with the CarHMM is that it assumes Markovian dynamics when conditioned on the hidden state, i.e. that any observation $Y^*_{t,t^*}$ depends only on the behavioral state $X^*_{t,t^*}$ and $Y^*_{t,t^*-1}$. However, there are many animal movement processes which violate this Markov property on very fine scales. For example, swimming behaviour of marine mammals often exhibits periodic behavior since the animal repeatedly flukes to propel itself forward. Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area (MAYBE?). With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine scale behavior is likely to persist.

To address this issue, we recommend borrowing techniques from signal processing literature to compress the data and summarize its essential elements. In particular, we suggest performing the discrete-time short-time Fourier transform (STFT) over the observed fine-scale process $Y^*_t$:
%
\begin{align*}
    STFT\{Y^*_{t,(s^*w)+1:(s^*+1)w) }\}(k) := \hat{Y}^{*(k)}_{t,s^*} = \sum_{n = 1}^{w} Y^*_{t,(s^*w)+n}e^{-i \frac{2\pi k}{w} (n-1)} \\ k = 0, 1, \ldots, w-1; \quad s^* = 0,1, \ldots, \lfloor T^*_t / w \rfloor - 1
\end{align*}
%
Where $i$ in the equation above refers the $\sqrt{-1}$. Note here that $s^*$ indexes the \textit{windows} moving across $Y^*_t$ while $t^*$ indexes the \textit{observations} of $Y^*_t$.

The STFT slides a moving window of length $w$ across the time series $Y_t^*$ and transforms the domain of each window from time to frequency. This allows the spectrum of $Y_t^*$ at time $t^*$ to be summarized by a $w$-dimensional vector of Fourier coefficients. While other step sizes can be used for the sliding window, we select $w$ to avoid serial dependence between windows.

If $Y^*_t \in \mathbb{R}^{T^*_t}$, then $\hat{Y}_t^* \in \mathbb{C}^{\lfloor T^*_t / w \rfloor \times w}$. Although this allows $Y^*_t$ to be represented in a way that eliminates obvious periodic behavior, the data set itself is still approximately as large as $Y^*_t$ itself. To reduce the size of $\hat{Y}^*_t$, we take two summary statistics of each window:
%
$$Z_{t,s^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,s^*}\right) \qquad Z_{t,s^*}^{*(2)} = \frac{1}{w}\sum_{k=1}^{\tilde{f}}|\hat{Y}^{*(k)}_{t,s^*}|^2$$
%
$Z_{t,s^*}^{*(1)}$ is equal to the average value of $Y_{t,s^*}^*$, and $Z_{t,s^*}^{*(2)}$ is equal to the squared 2-norm of the component of $Y_{t,s^*}^*$ that can be attributed to frequencies in the signal between $1$ and $\tilde{f}$ periods per window length. Both the window length $w$ and the max frequency $\tilde{f}$ are problem-specific tuning parameters. $w$ should be long enough to capture the periodic behavior of the underlying process (at least as long as the length of a period), but short enough to avoid over-smoothing of the data and to maintain high resolution in the behavioral process $X^*$. $\tilde{f}$ should be selected such that the maximum frequency of $Y_t^*$ that makes biological sense is $\tilde{f}$ per window length. Note that these summary statistics are just one possible choice to describe each window. Other choices of summary statistics include the maximum and argmax of $\hat Y^*_{s^*,t}$. Future studies can adjust these definitions as needed. A visualization of transforming a one-dimensional sequence $Y^*$ to $Z^*$ can be seen in figure (\ref{fig:fourier_example}).

\begin{figure}[h!]
	\centering
	\includegraphics[width=5in]{../Plots/fourier_transform.png}
	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and fourier transform.}
	\label{fig:fourier_example}
\end{figure}

@@

Fourier analysis has previously been used in the field of animal movement to explain animal behavior \cite{Fehlmann:2017} and specifically fluking \cite{Shorter:2017} from accelerometer data, and Fourier analysis within an HMM has been used to describe daily behavioral cycles of marine mammals  \cite{Heerah:2017}, but we are currently unaware of any studies which have incorporated Fourier analysis of accelerometer data within the structure of an HMM.

@@

Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes taking place simultaneously. However, as mentioned before, the fine-scale process $Y^*$ can often exhibit auto-correlation and intricate structure. Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit auto-correlation, especially in $Z_{t,s^*}^{*(1)}$. Therefore, we replace the fine-scale HMM within the Hierarchical HMM with a CarHMM, as shown in figure (\ref{fig:CarHHMM}).

\begin{figure}[h!]
	\centering
	\includegraphics[width=5in]{../Plots/CarHHMM.png}
	\caption{Graphical representation of a CarHHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
	\label{fig:CarHHMM}
\end{figure}

The likelihood of this model is still easy to calculate using the forward algorithm:
%
$$\calL_{\text{CarHHMM}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
%
where:
%
\begin{align*}
P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
&f^{(N)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
\end{align*}
%

>>>>

\subsection{Short-time Fourier Transform on Fine-scale processes}

One issue with the CarHMM is that it assumes Markovian dynamics when conditioned on the hidden state, i.e. that any observation $Y^*_{t,t^*}$ depends only on the behavioral state $X^*_{t,t^*}$ and $Y^*_{t,t^*-1}$. However, there are many animal movement processes which violate this Markov property on very fine scales. For example, swimming behaviour of marine mammals is often exhibits periodic since the animal repeatedly flukes (swims up and down) to propel itself forward. Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but to our knowledge addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area \textcolor{red}{(MAYBE?)}. With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine scale behavior is likely to persist.

We borrow techniques from signal processing literature to compress the data and summarize its essential elements. In particular, we suggest performing the discrete-time short-time Fourier transform (STFT) over the observed fine-scale process $Y^*_t$:
%
\begin{align*}
    STFT\{Y^*_{t,(s^*h+1):(s^*h+h) }\}(k) := \hat{Y}^{*(k)}_{t,s^*} = \sum_{n = 1}^{h} Y^*_{t,(s^*h)+n}e^{-i \frac{2\pi k}{h} (n-1)} \\ k = 0, 1, \ldots, h-1; \quad s^* = 0,1, \ldots, \lfloor T^*_t / h \rfloor - 1
\end{align*}
%
Where $i$ in the equation above refers the $\sqrt{-1}$. The STFT slides a moving window of length $h$ across the time series $Y_t^*$ and transforms the domain of each window from time to frequency. This allows the spectrum of $Y_t^*$ at time $t^*$ to be summarized by a $h$-dimensional vector of Fourier coefficients. While other step sizes can be used for the sliding window, we select $h$ to avoid serial dependence between windows. Note here that $s^*$ indexes the \textit{windows} moving across $Y^*_t$ while $t^*$ indexes the \textit{observations} of $Y^*_t$.

If $Y^*_t \in \mathbb{R}^{T^*_t}$, then $\hat{Y}_t^* \in \mathbb{C}^{\lfloor T^*_t / h \rfloor \times h}$. Although this allows $Y^*_t$ to be represented in a way that eliminates obvious periodic behavior, the data set itself is still approximately as large as $Y^*_t$ itself. To reduce the size of $\hat{Y}^*_t$, we take two summary statistics of each window:
%
$$Z_{t,s^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,s^*}\right) \qquad Z_{t,s^*}^{*(2)} = \frac{1}{h}\sum_{k=1}^{\tilde{f}}|\hat{Y}^{*(k)}_{t,s^*}|^2$$
%
$Z_{t,s^*}^{*(1)}$ is equal to the average value of $Y_{t,s^*}^*$, and $Z_{t,s^*}^{*(2)}$ is equal to the squared 2-norm of the component of $Y_{t,s^*}^*$ that can be attributed to frequencies in the signal between $1$ and $\tilde{f}$ periods per window length. Both the window length $h$ and the max frequency $\tilde{f}$ are problem-specific tuning parameters. $h$ should be long enough to capture the periodic behavior of the underlying process (at least as long as the length of a period), but short enough to avoid over-smoothing of the data and to maintain high resolution in the behavioral process $X^*$. $\tilde{f}$ should be selected such that $\tilde{f}$ periods per window length is the maximum frequency of $Y_t^*$ that makes biological sense. Note that these summary statistics are just one possible choice to describe each window, and other choices include the maximum and argmax of $\hat Y^*_{t,s^*}$. Future studies can adjust these definitions as needed. A visualization of transforming a one-dimensional sequence $Y^*$ to $Z^*$ can be seen in figure (\ref{fig:fourier_example}).

\begin{figure}[h!]
	\centering
	\includegraphics[width=5.5in]{../Plots/fourier_transform.png}
	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and Fourier transform.}
	\label{fig:fourier_example}
\end{figure}

@@

Fourier analysis has previously been used in the field of animal movement to explain animal behavior \cite{Fehlmann:2017} and specifically fluking \cite{Shorter:2017} from accelerometer data, and Fourier analysis within an HMM has been used to describe daily behavioral cycles of marine mammals  \cite{Heerah:2017}. However, we are currently unaware of any studies which have incorporated Fourier analysis of accelerometer data within the structure of an HMM.

@@

Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes. However, as mentioned before, the fine-scale process $Y^*$ can often exhibit auto-correlation and intricate structure. Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit auto-correlation, especially in $Z_{t,s^*}^{*(1)}$. Therefore, we replace the fine-scale HMM within the hierarchical HMM with a CarHMM as shown in (fig \ref{fig:CarHHMM}).
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=5in]{../Plots/CarHHMM.png}
	\caption{Graphical representation of a CarHHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
	\label{fig:CarHHMM}
\end{figure}
%
Naturally, we refer to this model as the \textbf{CarHHMM}, and it can be implemented both with and without the short-time Fourier-transform (STFT). The likelihood of this model is still easy to calculate using the forward algorithm:
%
$$\calL_{\text{CarHHMM}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
%
where:
%
\begin{align*}
P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
&f^{(N)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
\end{align*}
%
